---
title: "Disentangling a complex variable"
author: "Rob Kabacoff, Ph.D."
date: "March 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

## Untangling a complex variable

We are occassionally faced with a data set variable that contains multiple pieces of information. Consider the variable **charges** in the following dataset **test_data**.

```{r dataset, warning=F, message=F}
test_data <- data.frame(id=1:4,
                        charges = c("Murder,Kidnapping,Burglary",
                                    "burglary: assault and battery",
                                    "murder",
                                    ""))
knitr::kable(test_data, cpation="test_data")
```

Multiple pieces of information are stored in this variable. In this case these pieces are delineated by a comma or colon. As is, the variable is not useful.

The **untangle** function can be used to create a set of dummy codes from this variable that will be more useful. This results in a new dataset with dummy codes:

```{r function, warning=F}
untangle <- function(data, x, pattern, verbose=FALSE){
  require(stringr)
  
  variable <- str_to_lower(data[[x]])
  
  # obtain list of unique codes
  code_matrix <- str_split(variable, pattern, simplify=TRUE)
  code_vector <- as.character(code_matrix)
  code_vector <- str_trim(code_vector)
  code_vector <- unique(code_vector[code_vector != ""])
  if(verbose) cat("[Unique Codes] ", code_vector, "\n", sep="\n")
  
  # create dummy codes matrix
  nobs <- nrow(data)
  ncodes <- length(code_vector)
  dummy_codes <- matrix(rep(NA, times=nobs*ncodes), ncol=ncodes)

  # add 1/0 codes
  for(i in 1:ncodes){
    dummy_codes[,i] <- ifelse(str_detect(variable, code_vector[i]), 1, 0)
  }
  
  # add dummy code names
  dummy_codes <- as.data.frame(dummy_codes)
  codenames <- str_replace_all(code_vector, " ", "_")
  names(dummy_codes) <- codenames
  
  # add to data frame
  newdata <- cbind(data, dummy_codes)
  
}

```

```{r, echo=TRUE, warning=F}
test_data <- untangle(data=test_data, x="charges", pattern="[,:]", verbose=TRUE)

```



```{r results}
knitr::kable(test_data, caption="New test_data")

```


